---
title: "Life Expectancy Analysis"
output: html_document
date: "2024-05-14"
---

```{r setup, message=FALSE}
# Load standard libraries
library("tidyverse")
```
# Question 1

## Question 1.1

1. Life expectancy at birth refers to the number of years a newborn infant would live (or is expected to live) given that the existing mortality rates at the time of the birth remain the same throughout infant's life. In the case of period life expectancy, we  assume constant mortality rates into the future.

2. Done below. We renamed column 'time' to year. Removed the NA entries in lifeExpectancy and year columns (we found 1325 NAs in lifeExpectancy, 36 NAs in year, and none in name). Moreover, on inspection, we find some empty strings in name, iso2, and region columns. We will replace any empty strings ("") with NA and then remove them from our analysis data set. After data cleaning, we get 11618 'good' observations. 

```{r}
life_expectancy_df <- read.csv("gapminder.csv", sep="\t")
```

```{r}
life_expectancy_df <- life_expectancy_df %>% rename(year = time) #renamed column name time to year
```


```{r}
sapply(life_expectancy_df, class)
nrow(life_expectancy_df) 
ncol(life_expectancy_df)
sum(is.na(life_expectancy_df$lifeExpectancy))
sum(is.na(life_expectancy_df$year))
sum(is.na(life_expectancy_df$name))
```

```{r}
life_expectancy_df <- life_expectancy_df %>% drop_na(lifeExpectancy)
life_expectancy_df <- life_expectancy_df %>% drop_na(year)
```

As seen below using the function summary, values in our dataset look reasonable.

```{r}
summary(life_expectancy_df$lifeExpectancy)
summary(life_expectancy_df$year)
```

```{r}
life_expectancy_df[life_expectancy_df == ""] <- NA 
life_expectancy_df <- life_expectancy_df %>% drop_na(name)
```

3a. 204 countries.
```{r}
countries <- unique(life_expectancy_df$name)
length(countries)
```

3b. first year is 1960 and last or latest year is 2019.
```{r}
min(life_expectancy_df$year)
max(life_expectancy_df$year)
```

3c. Lowest life expectancy is 18.907 yrs. It was recorded in Cambodia in 1977. Highest recorded life expectancy is 85.42 yrs for San Marino in Europe in the year 2012.

3d. The Cambodian Genocide in the late 1970s was responsible for mass violence and the resultant low life expectancy. It is estimated that about 1.5 to 3 million people were killed by Khmer Rouge, a communist political group.
```{r}
life_expectancy_df %>% filter(lifeExpectancy == min(lifeExpectancy))
life_expectancy_df %>% filter(lifeExpectancy == max(lifeExpectancy))
```

4. Shown below. We have selected Rwanda as it shows an interesting drop in the life expectancy in early 1990s. We have also selected Afghanistan and India as both these countries had lower than average life expectancy in 1960 but their consistent positive growth will be insightful to compare with that of developed and under-developed countries. Moreover, we select Srilanka and try to compare how neighbors India and Srilanka have performed knowing that Srilanka started higher than India in the 1960. 

```{r}
life_expectancy_df %>% group_by(name) %>% summarize(lowest = min(lifeExpectancy)) %>% arrange(lowest) %>% head()

#Rwanda sticks out and is selected
```

```{r}
life_expectancy_df_selected <- life_expectancy_df %>% filter(name %in% c("United States of America", "Korea, Republic of", "Cambodia", "China", "Rwanda", "Afghanistan", "India", "Sri Lanka"))
```

```{r}
ggplot()+
  geom_line(mapping = aes(x = year, y = lifeExpectancy, group=name), alpha=0.4, color = "gray", data=life_expectancy_df)+
  geom_line(mapping = aes(x = year, y = lifeExpectancy, group=name, color=name), data= life_expectancy_df_selected)+
  ggtitle("Life Expectancy vs Years")+
  xlab("Year")+
  ylab("Life Expectancy")+
  theme(plot.title = element_text(hjust = 0.5))+
  labs(color = "Country")
```

5. A very high level view tells us that life expectancy increases over our time range (1960-2019). However, the rate of growth is different for different countries, and a few countries in our selection and the overall dataset go through interesting peaks and troughs in this time range. Firstly, life expectancy in USA has changed by 12.93% from 69.77 to 78.79 yrs. This trend showcases a steady increase in the US life expectancy over the years. Cambodia, on the other hand, had the lowest life expectancy in 1977 and the huge golden trough showcases the same. However, it is interesting to note that Cambodia has mostly overcome the 1970s genocide and latest life expectancy is ~70 yrs. Rwanda's life expectancy dropped in the early 1990s due to systemic massacres which were triggered after the deaths of Presidents of Rwanda and Burundi in a plane crash caused by a rocket attack. 

China has taken giant leaps in life expectancy growth over the time period. It records a staggering 75.91% increase from 1960-2019 owing to robust healthcare development and administration by the government. South Korea's life expectancy has grown by 50.2% to ~83.23 yrs. There are several interesting theories behind East Asia's high life expectancy numbers. One suggests that the food and dietary habits are excellent, providing holistic nutrition.

**comparison of the Neighbors**: India started at a low life expectancy of 41.42 yrs in 1960. Although the country has seen bouts of political instability, such as the suspicious death of Prime Minister Lal Bahadur Shastri in 1966 and the 1984 anti Sikh riots, life expectancy has steadily climbed over the years owing to general stability and quick damage control actions by the government agencies. From 1960-2019, India registered a life expectancy growth of 68.18%. On the other hand, Srilanka, a much smaller country in terms of GDP output and area, has shown remarkable performance as it has a life expectancy growth rate of 29.66% with a 2019 life expectancy of 76.98 yrs! Ignoring the small negative effect of Srilankan civil war from 1980-2000 (this affect is visible in the visualization), Srilanka's excellent healthcare and immunization programs have been responsible for a very high life expectancy.

6. Done below.
```{r}
#dropping NA entries in the fertility rate column
life_expectancy_df <- life_expectancy_df %>% drop_na(fertilityRate)
```

```{r}
life_expectancy_df_selected <- life_expectancy_df %>% filter(name %in% c("United States of America", "Korea, Republic of", "Cambodia", "China", "Rwanda", "Afghanistan", "India", "Sri Lanka")) #updating selected data frame
```

```{r, warning=FALSE}
#first plot is all countries in semi transparent gray
first_plot <- ggplot()+
  geom_point((mapping = aes (x=fertilityRate, y = lifeExpectancy, group=name)), color="gray", alpha = 0.4, size = 1, data=life_expectancy_df)

#using arrow option in geom segment to show trend. function 'lead' ensures arrows point towards the next data point 
first_plot +
  geom_segment(mapping = aes(x = fertilityRate, y = lifeExpectancy, xend = lead(fertilityRate), yend = lead(lifeExpectancy), group=name), color = "gray", alpha=0.4, arrow=arrow(length = unit(0.07, "inches")), data=life_expectancy_df) + 
  geom_segment(mapping = aes(x = fertilityRate, y = lifeExpectancy, xend = lead(fertilityRate), yend = lead(lifeExpectancy), group=name, color=name), arrow=arrow(length = unit(0.07, "inches")), data=life_expectancy_df_selected)+
  ggtitle("Life Expectancy vs Fertility Rate")+
  xlab("Fertility Rate")+
  ylab("Life Expectancy")+
  theme(plot.title = element_text(hjust = 0.5))+
  labs(color = "Country")
```

7. A high level observation is that world is going towards decreasing fertility rate with an increasing life expectancy. This result is also in sync with the expectation that as countries develop and prosper over time, healthcare infrastructure improves, in-turn resulting in higher awareness around family planning. In other words, 'quality' gets precedence over 'quantity'. For US, China, South Korea and Cambodia, fertility rates have decreased over 1960-2019. The most noticeable downward trend seems to be in China and Korea wherein the fertility rate has decreased from around 6 to ~0.92 for Korea and ~1.7 for China. US has had a 53% drop in fertility rate from 3.65 in 1960 to 1.7 in 2019.

In India, the fertility rate has dropped to ~2.2 in 2019 from around 5.9 yet the decrease is not as sharp as some of the other East Asian nations such as South Korea or China. Interestingly, Afghanistan had a high fertility rate of 7.45 in 1960 and while it has gone down, it still is around ~4.32, indicating that the country is currently evolving and working on healthcare accessibility and awareness campaigns.

All in all, it's not just socioeconomic factors such as economic development and healthcare accessibility but also reasons such as women empowerment and changing cultural norms which are responsible for an inverse relationship between fertility rate and life expectancy.

## Question 1.2

1. The distribution of life expectancy is a left-skewed distribution. In order to test if it is log normal, we do a log transformation and don't get a normal distribution, meaning it is not log normal and hence a log transformation is not helpful. We will proceed as is without any transformation.

```{r}
hist(life_expectancy_df$lifeExpectancy, breaks=150)
life_expectancy_df$log_le <- log(life_expectancy_df$lifeExpectancy)
hist(life_expectancy_df$log_le, breaks=150)
```


```{r}
life_expectancy_df <- life_expectancy_df %>% mutate(log_le = NULL) #deleting the log transformed column
```

2. Done below.
```{r}
life_expectancy_df$yearminus2k <- life_expectancy_df$year - 2000
```

```{r}
linear_m <- lm(lifeExpectancy ~ yearminus2k, data = life_expectancy_df)
summary(linear_m)
```

3. year - 2000 makes sense because it will help us reasonably understand the meaning behind intercept estimate. With a y - 2000 transformation, we can say that the intercept estimate is life expectancy for the year 2000 because (year - 2000) is zero when year is 2000. The transformation standardizes the time scale and now has a meaningful central point of year = 2000. If we do not do the transformation, we are left to make sense out of year = 0, which doesn't have any intuitive meaning.

4. B0 (67.34 yrs) is the life expectancy for the year 2000. B1 is ~0.309. It refers to the change in life expectancy when we add 1 year. Life expectancy increases by 0.309 yrs over every year. Results are statistically significant as p value is <2e-16.

```{r}
sum(is.na(life_expectancy_df$region)) #shows no missing values in region
```

```{r}
unique(life_expectancy_df$region) #result shows we have 5 regions overall
```

5. Done below.
```{r}
multiple_m <- lm(lifeExpectancy ~ yearminus2k + region, data = life_expectancy_df)
summary(multiple_m)
```

6. Africa is the reference category. Intercept estimate is 55.87 yrs, which is the life expectancy for the year 2000 and region Africa. The life expectancy for America is 55.87 + 15.93 = 71.8 yrs, keeping the year same. Likewise, Asia's life expectancy is 12.22 yrs more than that of Africa. Oceania's is 13.65 yrs more than the reference category's or Africa's. Europe has the highest life expectancy, which is ~20.9 yrs more than Africa's. For all these comparison statement, year or time is kept the same. All results are statistically significant since p value is very low (<2e-16).

Lastly, if we increase year by 1 without changing other variables, life expectancy increases by ~0.305 yrs. Result is statistically significant since p value is <2e-16 (very low!). Magnitude is slightly smaller than what we previously saw.

```{r}
life_expectancy_df <- life_expectancy_df %>% drop_na(GDP_PC) #dropping NA entries
life_expectancy_df$log_GDPPC <- log(life_expectancy_df$GDP_PC)
```

7. Shown below.
```{r}
multiple_m_v2 <- lm(lifeExpectancy ~ yearminus2k + region + log_GDPPC + fertilityRate, data = life_expectancy_df)
summary(multiple_m_v2)
```
8. The log_GDPPC estimate is 2.49, which tells us that life expectancy increases by 2.49 yrs when we increase log of GDP per capita by 1 unit. Also, as the fertility rate increases by 1 unit, life expectancy decreases by 2.23 yrs. It is key to note that we're assuming other variables stay the same. All results are statistically significant.

9. America is now leading the pack with a dummy estimate of 6.034. Europe's estimate is 4th largest - 5.42. Earlier, we modeled life expectancy on time and region and now we are trying to understand how log GDP per capita and fertility rate along with the other two explanatory variables affect life expectancy. Now there is an interplay of several more variables in our model, and we are interested in understanding the cumulative affect of all these variables on our outcome variable, life expectancy. Our new model adjusts Beta values according to all the explanatory variables and hence the rankings change.

```{r}
life_expectancy_df %>% group_by(region) %>% summarize(avg_life_expectancy=mean(lifeExpectancy)) #checking average life expectancy
```
10. Basis our final model which gave statistically significant results, we can say America has the highest life expectancy. Our final model which says America has the highest life expectancy has a R square of 0.847, best among all our models. However, as per group averages Europe leads the pack and Europe also comes out on the top when we modeled life expectancy on only year and region. Africa is certainly the lowest and the result is confirmed in every model. 
