1. Done below. No missing values. However, we need to convert East column into a factor variable.

```{r}
restaurants_df <- read.csv("nyc-italian-cheap.csv", sep="\t")
```

```{r}
nrow(restaurants_df)
ncol(restaurants_df)
sapply(restaurants_df, class)
sum(is.na(restaurants_df))
restaurants_df[restaurants_df == ""] <- NA 
sum(is.na(restaurants_df))
summary(restaurants_df)
```

2. Because the outcome variable is a binary variable and can have two and only two distinct values 0 and 1, we will model the probability using logistic regression. In other words, since we are trying to predict whether a restaurant is cheap or not, there are only two associated values for the outcome variable. Cheap is 1 and not cheap or expensive is 0. Logistic is best suited for the same reason.

We convert East into a factor variable. Numeric operations do not make sense on 'East' as it refers to a location based category.

```{r}
restaurants_df$East <- as.factor(restaurants_df$East)
class(restaurants_df$East)
```

3. Done below. Explanation follows the code.

```{r}
log_m <- glm(Cheap ~ Food + Decor + Service + East, data = restaurants_df, family=binomial())
summary(log_m)
library(margins)
me <- margins(log_m)
summary(me)
```

We have assumed 4 explanatory variables - Food, Service, Decor and East/location. Since we performed logistic regression, we will use the marginal effect to explain the results. Also, as AME is the Average Marginal Effect, it's necessary to keep in mind that these estimates are all 'averages' because the probability will vary unequally (logistic function is not linear) for an increase of 1 unit in the explanatory variable. 

Firstly, we can say that increasing the decor review by 1 unit decreases the probability of the restaurant being cheap by 8.5% on an average. Similarly, if we increase the food review by 1 unit, the probability of cheap restaurant goes down by 6.37% on an average. Both the above results are statistically significant as the p value is low enough, hence we can reject the corresponding null hypothesis. 

Moreover, we have two statistically insignificant results. Their insignificance is confirmed by the high p value (0.45 and 0.8). One is that for a restaurant located east of the 5th Ave, probability of it being cheap goes down by 4.49%. Second is if we increase service review by 1 unit, probability of that restaurant being cheap increases by 0.64%. The latter is non intuitive and statistically insignificant. We can discard both from our results.

I believe the overall results align with common sense. Given that we increase decor in any restaurant, it is likely an expensive restaurant and hence the Probability of finding a cheap restaurant diminishes. Expensive restaurants like to work on ambiance to provide an overall customer experience and do not limit themselves to prioritizing just delicious food. Also, it is probable that cheaper restaurants compromise on key aspects such as food quality to cut costs, prompting customers to give poor 'food' reviews. Hence, our model states that improving food reviews decreases the probability that the restaurant is cheap.


4. Done below. Analysis follows the code.
```{r}
#creating a data frame for our different restaurant data
food <- c(23, 18)
decor <- c(17, 15)
service <- c(22,24)
east <- c(0,1)
```

```{r}
newdata <- data.frame(Food=food, Decor=decor, Service=service, East=east)
newdata$East <- as.factor(newdata$East) #converting East into a factor variable
predict(log_m, newdata, type="response")
```

```{r}
predict(log_m, newdata, type="response") > 0.5
```

First we create a dataframe 'newdata' to store review data for the two new restaurants. Then, we convert East into a factor variable and use predict() to ask our model to work on this new dataset and generate probability whether the restaurant is cheap or not. 

The probability that the first restaurant (Assagio Ristorante) is cheap is ~0.236, so we can classify it as an expensive restaurant. For Altura, this probability is 0.902 and we classify it as a cheap restaurant.

Conclusion - Our model predicts that Altura is a cheap restaurant while Assagio Ristorante is not. 
